Heavy rainfall in Paraguay during the 2015-2016 austral summer: causes and sub-seasonal-to-seasonal predictive skill

### 2018-03-03 reports/writeup/paper.tex
james writing edits

@@

\title{Heavy rainfall in Paraguay during the 2015-2016 austral summer: causes and subseasonal-to-seasonal predictive skill}
\authors{James Doss-Gollin\correspondingauthor{James Doss-Gollin. Columbia Water Center. Dept. of Earth and Environmental Engineering, Columbia University, 500 W. 120th St., New York, NY. USA.}}
\affiliation{Columbia Water Center. Dept. of Earth and Environmental Engineering, Columbia University, 500 W. 120th St., New York, NY. USA \\ Columbia Water Center, Columbia University, 500 W. 120th St., New York, NY. USA}
\email{james.doss-gollin@columbia.edu}
\extraauthor{\'Angel G. Mu\~noz}
\extraaffil{Atmospheric and Oceanic Sciences (AOS). Princeton University. Princeton, NJ 08540-6649. USA  \\
	International Research Institute for Climate and Society (IRI). The Earth Institute. Columbia University. Palisades, NY. 10964-1000. USA}
\extraauthor{Simon J. Mason}
\extraaffil{International Research Institute for Climate and Society (IRI). The Earth Institute. Columbia University. Palisades, NY. 10964-1000. USA}
\extraauthor{Max Past\'{e}n}
\extraaffil{Direcci\'{o}n de Meteorolog\'{i}a e Hidrolog\'{i}a. Paraguay \\ Facultad Polit\'ecnica. Universidad Nacional de Asunci\'{o}n. Paraguay}

@@

During the austral summer of 2015-16, repeated heavy rainfall events led to severe flooding in the Paraguay River basin, displacing approximately \num{170000} people \citep{BrackenridgeDFO} and causing tremendous damage to property and infrastructure \citep{MinisteriodeObrasPublicasyComunicacion:2016tq}.
Because population in South America tends to concentrate along coasts and rivers (see supplemental figure S1), flooding in the Paraguay River basin directly affects not only much of the population of Paraguay, but also of populations in Argentina and Uruguay who lie along the Paran\'{a} and la Plata rivers into which the Paraguay River drains.
Heavy rainfall and flooding in the basin also has important implications for hydropower generation, for agriculture, and for regional water resource management.
The aim of this paper is to diagnose the drivers of the November-February (NDJF) 2015-16 rainfall and flooding events, and to assess the skill of the associated subseasonal-to-seasonal predictions.

The rainfall climatology of the Paraguay River basin varies strongly by season, with extratropical characteristics in the winter and monsoonal characteristics in the summer.
In the warm season (NDJF) which is the focus of this study, circulation is dominated by the upper-tropospheric Bolivian High, the lower-level subtropical highs, the Chaco Low over northern Argentina, the South Atlantic Convergence Zone (SACZ), and the South American Low-Level Jet, or SALLJ \citep{Grimm2009,Marengo:2012cm}.
Rainfall peaks around \SI{5}{\milli\meter\per\day} during the warm months (October-May) and reaches a minimum near \SI{2}{\milli\meter\per\day} in July and August.
However, the flat topography limits the river's ability to carry the summer runoff, causing seasonal inundation of the Pantanal and distributing the river discharge in time \citep{Bravo:2011et,Barros:2004bn}.
Thus, the streamflow maxima typically occurs in phase with precipitation upstream of the Pantanal, while downstream of the Pantanal -- which we define as the Lower Paraguay River Basin; see \cref{fig:study-area} -- the annual peak typically occurs between April and July.

During the warm season (NDJF) a large fraction of rainfall, and nearly all heavy rainfall, in the Paraguay River Basin is associated with mesoscale convection \citep{Velasco1987}.
Previous studies of organized convection and precipitation across subtropical continental South America have found close correspondence to the exit region of the low-level jets \citep{Saulo:2007km,Salio:2007gd,Marengo2004,Velasco1987}, which is influenced in both summer and winter by midlatitude baroclinic wave trains that interact with the Andes topography to generate orographically bound cyclones and northerly low-level flow \citep{Campetella:2002hx,Seluchi:2006bi,Boers:2014gt}.
The strength and direction of this moisture transport varies substantially between events; SALLJ exit regions range from central Argentina \citep[``Chaco Jet Events'';][]{Salio:2002ev} to Paraguay and southeastern Brazil \citep[``No-Chaco Jet Events'';][]{Vera:2006ib}.

At sub-seasonal timescales, convection in the Paraguay River Basin is modulated by a variety of drivers, notably including the SACZ and the Madden-Julien Oscillation (MJO).
During SACZ conditions, strong convergence is observed over the Amazon basin with divergence over southwestern Brazil, northern Argentina and Paraguay \citep{Herdies:2002jy,Carvalho2010}; the opposite is true for so-called NoSACZ conditions.
SACZ occurrence is related to westerly wind regimes, as well as ``active'' and ``break'' periods of the South American Monsoon System \citep{Marengo2004}.

At seasonal timescales, El Ni\~no-Southern Oscillation (ENSO) is the dominant driver of convection variability in the Paraguay River Basin.
During El Ni\~no years, a low-level anticyclonic anomaly over central Brazil enhances occurrence of the low-level jet, favoring the development of mesoscale convective systems \citep{Velasco1987}.
The intensity and precise extent of this anomaly is relevant for the precise impact of ENSO events.
The region also exhibits substantial variability between seasons of rainfall during El Ni\~no years, including a reversal of rainfall anomalies between November of that year and January of the following one, influenced by land-surface interactions \citep{Grimm2009,Grimm2003}.
Even beyond El Ni\~no years, regional land-surface feedbacks can cause regions that exhibit wet anomalies in the spring to experience more summer precipitation on average \citep{Grimm:2009bq}.
Similarly, mid-latitude dynamics influence low-level wind anomalies on many time scales, though this relationship is complicated due to coupled tropical-extratropical interactions \citep{Carvalho2004,Jones2002}.
To address these potential interactions, a cross-timescale approach based on synoptic circulation types is employed here to diagnose the causes of the rainfall events; this method has been used in previous work for southeastern South America \citep{Munoz2015,Munoz2016} and other regions \citep{Moron2015}.

@@

Rainfall data are taken from the CPC Unified Gauge-based Analysis of Global Daily Precipitation dataset \citep{xie2010cpc}.
Spatial resolution is \SI{0.5}{\degree}, and temporal resolution is daily.
The retro and real-time datasets are combined and treated as a single dataset.
We define ``heavy'' rainfall events to be exceedances of the \nth{90} percentile; while the value is different for each grid cell, the \nth{90} percentile of area-averaged rainfall over the Lower Paraguay River Basin is approximately \SI{15}{\milli\meter\per\day}.

Atmospheric circulations are diagnosed using daily data from the NCAR-NCEP Reanalysis II dataset \citep{Kanamitsu:2002kk}.
Spatial resolution is \SI{2.5}{\degree}.
Because the end-of-day time for the rainfall data is 12:00GMT over most of South America \citep{xie2010cpc}, we use six-hour reanalysis data, and shift by twelve hours before re-sampling to the daily time step.
This ensures that the time steps in the reanalysis and rainfall data sets are the same, but means that a day is defined as beginning at 12:00 GMT.
Since most summer rainfall in this region occurs overnight, this end-of-day time (which translates to approximately 8:00 AM locally depending on the exact time zone) tends to separate distinct events.
Ocean sea surface temperature (SST) patterns are explored at the monthly time step using the \SI{1}{\degree} NOAA OI.v2 dataset \citep{Reynolds:2002iy}.

Streamflow data was collected by the Paraguayan Navy and National Administration of Navigation and Ports of Paraguay, and was processed and distributed by the Paraguayan Directorate of Meteorology and Hydrology (DINAC-DMH).
Locations of streamflow gauges are shown in \cref{fig:study-area}.
Because no stage-discharge curves are available, we present only the river stage values; while this is relevant from the perspective of flood damage, flow rates cannot be estimated without these curves (which are difficult to reconstruct as river geometry changes over time).

This study also makes use of some climate indices.
Data on the El Ni\~{n}o-Southern Oscillation (ENSO), or NI\~NO3.4, came from a statistical-dynamical interpolation \citep{Kaplan:1998df}, which is constrained by relatively high-quality observations during the study period.
Data on the Madden-Julien Oscillation (MJO) came from the Australian Bureau of Meteorology \citep{Wheeler:2004ea}.

@@

The seasonal predictions used were ``flexible format'' forecasts provided by the International Research Institute for Climate and Society (IRI).
This product provides the probability of exceedance (or non-exceedance) of a user-specified percentile of the climatological distribution.
The DJF 2015-2016 forecasts analyzed were produced in November 2015.
Due to the short record of flexible format forecasts (2012-2016), no verification was performed for seasonal predictions.
These forecasts are provided at a horizontal resolution of \SI{2.5}{\degree}.

The sub-seasonal forecasts used were issued by the European Centre for Medium-Range Weather Forecasts (ECMWF) using a coupled model, available via the Subseasonal-to-Seasonal Prediction Project Database \citep{Vitart2016} at \SI{1.5}{\degree} resolution.
Forecasts consider the period starting in Dec 2015 until Mar 2016 and hindcasts to assess the real-time predictive skill consider the period Dec 1-7, 1995-2014.
There is a total of 51 ensemble members for each forecast, and 11 ensemble members for each of the 20 hindcasts (Dec 1-7, 1995-2014).

@@

Anomalies were calculated relative to the seasonal mean from November 1979 to February 2016, thus retaining the seasonal cycle.

@@

Several types of analyses are used to diagnose the causes of the heavy rainfall events, and to bias-correct and verify the forecasts.
All codes used for data and fully reproducible analysis are available at \url{https://github.com/jdossgollin/PYFloods}.
Computation was performed in the python environment using the \texttt{xarray} \citep{hoyer2017xarray} and \texttt{numpy} \citep{vanderWalt:2011dp} packages for gridded data and the \texttt{pandas} \citep{McKinney:2010un} package for tabular data; plotting was performed with the \texttt{matplotlib} \citep{Hunter:2007ux} and \texttt{cartopy} \citep{Cartopy} packages.

Given the behavior of the Paraguay River discussed above, we define the Lower Paraguay River Basin as the region bounded by \SIrange{-59.75}{-55.75}{\degree W} and \SIrange{26.75}{22.75}{\degree S}, as shown in \cref{fig:study-area}.
This region corresponds to the region over which, given topography and previous studies \citep{Barros:2004bn,Bravo:2011et}, one might hypothesize rainfall inputs to most closely correspond to river levels at the stream gauges in \cref{fig:study-area}.
Results are not sensitive to small variations in this domain.

@@

The clustering was performed on the daily NDJF \SI{850}{\hecto\pascal} streamfunction field ($\psi$), calculated by integrating the meridional and zonal wind fields using spherical harmonics, as implemented in the \texttt{windspharm} package \citep{Dawson:2016ws,Adams:1999fb}, over the domain spanning \SIrange{15}{30}{\degree S} and \SIrange{65}{45}{\degree W} (\cref{fig:study-area}).

To facilitate clustering (which tends to perform poorly in high-dimensional spaces), the NDJF anomaly field of $\psi_{850}$ was projected  onto its four leading empirical orthogonal functions (EOFs), accounting for $>95\%$ of the total observed variance (supplemental figure S2), using the \texttt{eofs} package \citep{Dawson:2016ge}.
No meridional weighting was applied as the selected domain is relatively small and does not extend into high latitudes.
Once the EOFs were calculated, the principal component time series were computed for each day and scaled to unit variance.
This rescaling is not a necessary step; its effect is to treat all retained principal components as equally important for the weather typing, which provides relatively greater weight to EOFs 2, 3, and 4 than carrying out weather typing without re-scaling.
Though our approach of first selecting the number of EOFs to use and then choosing to scale them equally involves more subjective decisions than an approach without rescaling, in this case the resulting physical mechanisms are more interpretable.

Next, the $K$-means algorithm was used to assign a single cluster value to each day on record using the 4-dimension principal component time series.
The $K$-means technique is a partitioning method that classifies all days in the study into a predefined number of clusters.
The algorithm proceeds as follows:
\begin{enumerate}
	\item Randomly choose $K$ cluster centers $\mu_1^{(0)}, \ldots, \mu_K^{(0)}$ (where $0$ refers to the $0$th iteration)
	\item Iterate until convergence, indexing each iteration with $j$:
	\begin{enumerate}
		\item Assign each observation (day) $x_i$ to the nearest cluster center; we define this using the Euclidean distance but other measures, such as the Mahalanobis distance, could also be used:
		\begin{equation}
			m_i^{(j+1)} := \arg\min_{k\in 1,\ldots,K} \abs{\abs{x_i - \mu_k^{(j)}}}
		\end{equation}
		\item Recompute the cluster centers as the mean of all points assigned to that cluster
		\begin{equation}
			\mu_k^{(j+1)} := \frac{1}{\abs{\qty{i \big| m_i^{(j+1)}=k}}} \sum_{i | m_i^{(j+1)}=k} x_i
		\end{equation}
		where $\abs{\cdot}$ denotes vector length.
		\item Stop iteration if the change in centroids $\mu^{(j+1)} - \mu^{(j)}$ is less than a small but non-zero tolerance parameter $\tau$.
	\end{enumerate}
\end{enumerate}
The cluster centroids (weather types) $\mu_k$ produced by the $K$-means algorithm can then be interpreted as a Voronoi decomposition of the phase space into $K$ regions, and specifically as the Voronoi diagram which minimizes within-cluster variance.

The $K$-means algorithm is guaranteed to converge to a local minimum of intercluster variance; to select the best partition, 500 simulations were created using the implementation in Python's \texttt{scikit-learn} package \citep{Pedregosa:2012tv} and then the classifiability index of \citet{Michelangeli1995} was computed between each partition and the 499 others.
The partition whose classifiability index, averaged for all 499 pairwise comparisons, was the highest was selected.
Calculation of the classifiability index for several values of $K$ (supplemental figure S3) suggests that states with $K=5,6,\ldots,9$ are all reasonable; we chose the solution $K=6$ because the weather types identified are qualitatively similar to those determined over southeastern South America \citep{Munoz2015,Munoz2016} and have an intuitive physical meaning, discussed further in the following sections.

\subsection{Forecasts and Model Output Statistics (MOS)}

A wide variety of methods, generically known as Model Output Statistics \citep[MOS;][]{Glahn:1972vt}, have been proposed to correct for different types of bias in model outputs.
In this work, we analyze how well the rainfall events could have been predicted, both using the raw sub-seasonal forecasts and  MOS-adjusted sub-seasonal forecasts.
We use  four types of MOS techniques: the homoscedastic and heteroscedastic versions of the extended logistic regression (XLR and HXLR, respectively); principal component regression (PCR); and canonical correlation analysis (CCA).

Logistic regression models the conditional probability of binary events, and it has been used in MOS approaches for binary predictands \citep{Hamill:2004hk}.
Nonetheless, when using the logistic regression to address multiple thresholds via independent fits, the predicted probabilities are, in general, not mutually consistent \citep{Messner:2014gp}.
The (homoscedastic) XLR was designed to address this shortcoming \citep{Wilks:2009bk}, via the consideration of a transformation of the thresholds of interest as an additional predictor variable.
The HXLR, a heteroscedastic generalization of the XLR, was proposed to appropriately use the ensemble spread as predictor for the dispersion of the predictive distribution \citep{Messner:2014gp}.

CCA is a common statistical method frequently used to forecast rainfall using a purely empirical approach \citep{Mason:2008da,Barnston:2012ce,Jolliffe2012:vm,Barnston:1992gd,Wilks:2006fx}.
CCA identifies modes of co-variability, called canonical variates or canonical modes, by maximizing the correlation between linear combinations of the predictor and predictand's empirical orthogonal functions (EOFs).
The method forecasts spatial patterns of variability spanning across the region of interest rather than making forecasts for individual locations.
In PCR, a special case of CCA, each grid cell in the predictand field is estimated by regression using a linear combination of the predictor's EOFs \citep{Mason:2008da,Wilks:2006fx} rather than by identifying canonical modes.
Unlike the XLR and HXLR models, which perform bias correction independently for each grid cell, the CCA and PCR models can address biases in both the magnitude and the spatial distribution of the modeled precipitation patterns.

For the purposes of MOS corrections, the predictand (variable to forecast) is the observed rainfall for the target period of interest, and the predictor (variable to be corrected) is the raw model rainfall for the same period.
Exceedance of the \nth{90} percentile during the 1995-2014 period is used to define the heavy event cases.
We use the same spatial domain [\SIrange{39}{17}{\degree S}; \SIrange{66}{49}{\degree W}] for both the predictor and the predictand, except for the PCR and CCA cases, in which a larger domain [\SIrange{0}{60}{\degree S}; \SIrange{80}{30}{\degree W}] was used to better capture the spatial patterns in the raw model field.
A variety of domains and ways to combine initialization times were explored; the best results were selected in terms of the corresponding Kendall's $\tau$ rank correlation coefficient.
A summary of the final candidate predictors found to be most skillful for each MOS model is presented in \cref{tab:mos-methods}.

@@

The IRI's seasonal predictions are already provided with bias correction \citep{Barnston:2010ge}, and thus we did not perform any further MOS on seasonal rainfall fields.

@@

In addition to visually comparing predictions and observations to verify how well the heavy rainfall events could have been predicted, we use the Ignorance Score,
\begin{equation}\label{eq:ignorance}
	\text{IGN} \equiv - \log_2 p(Y),
\end{equation}
where $Y$ is the observed outcome and $p(Y)$ is the density function of the forecast distribution \citep{Good:1952ux,Roulston:2002jz,Brocker:2007er}.
The Ignorance score was introduced as the fundamental information-based verification measure,  decomposable into easily interpretable components: reliability, resolution and uncertainty \citep{Weijs:2010cl}.
Due to its close relationship to Shannon's information entropy, it is used to measure forecast utility, or the amount of information gain expected from a forecast \citep{Roulston:2002jz}.

>>>>

\usepackage[acronym]{glossaries}
\makeglossaries
\newacronym{wt}{WT}{weather type}
\newacronym{sallj}{SALLJ}{South American low-level jet}
\newacronym{enso}{ENSO}{El Ni\~{n}o-Southern Oscillation}
\newacronym{mjo}{MJO}{Madden-Julian Oscillation}
\newacronym{sacz}{SACZ}{South Atlantic Convergence Zone}
\newacronym{iri}{IRI}{International Research Institute for Climate and Society}
\newacronym{ecmwf}{ECMWF}{European Centre for Medium-Range Weather Forecasts}
\newacronym{s2s}{S2S}{sub-seasonal to seasonal}
\newacronym{eof}{EOF}{empirical orthogonal function}
\newacronym{mos}{MOS}{model output statistics}
\newacronym{xlr}{XLR}{homoscedastic extended logistic regression}
\newacronym{hxlr}{HXLR}{heteroscedastic extended logistic regression}
\newacronym{pcr}{PCR}{princiap components regression}
\newacronym{cca}{CCA}{canonical correlation analysis}
\newacronym{sst}{SST}{sea surface temperature}

@@

\title{Heavy rainfall in Paraguay during the 2015-2016 austral summer: causes and sub-seasonal-to-seasonal predictive skill}
\authors{James Doss-Gollin\correspondingauthor{James Doss-Gollin. Columbia Water Center. Dept. of Earth and Environmental Engineering, Columbia University, 500 W. 120th St., New York, NY. USA.}}
\affiliation{Columbia Water Center. Dept. of Earth and Environmental Engineering, Columbia University, 500 W. 120th St., New York, NY. USA \\ Columbia Water Center, Columbia University, 500 W. 120th St., New York, NY. USA}
\email{james.doss-gollin@columbia.edu}
\extraauthor{\'Angel G. Mu\~noz}
\extraaffil{Atmospheric and Oceanic Sciences (AOS). Princeton University. Princeton, NJ 08540-6649. USA  \\
	International Research Institute for Climate and Society (IRI). The Earth Institute. Columbia University. Palisades, NY. 10964-1000. USA}
\extraauthor{Simon J. Mason}
\extraaffil{International Research Institute for Climate and Society (IRI). The Earth Institute. Columbia University. Palisades, NY. 10964-1000. USA}
\extraauthor{Max Past\'{e}n}
\extraaffil{Direcci\'{o}n de Meteorolog\'{i}a e Hidrolog\'{i}a. Paraguay \\ Facultad Polit\'ecnica. Universidad Nacional de Asunci\'{o}n. Paraguay}

@@

During the austral summer of 2015-16, repeated heavy rainfall events led to severe flooding in the Paraguay River basin, displacing approximately \num{170000} people \citep{Brakenridge2016} and causing tremendous damage to property and infrastructure \citep{MinisteriodeObrasPublicasyComunicacion2016}.
Because population in South America tends to concentrate along coasts and rivers (see supplemental figure S1)\footnote{check}, flooding in the Paraguay River basin directly affects not only much of the population of Paraguay, but also of populations in Argentina and Uruguay who lie along the Paran\'{a} and la Plata rivers into which the Paraguay River drains.
Heavy rainfall and flooding in the basin also has important implications for hydropower generation, for agriculture, and for regional water resource management.
The aim of this paper is to diagnose the drivers of the November-February (NDJF) 2015-16 rainfall and flooding events, and to assess the skill of the associated subseasonal-to-seasonal predictions.

The rainfall climatology of the Paraguay River basin varies strongly by season, with extratropical characteristics in the winter and monsoonal characteristics in the summer.
In the warm season (NDJF) which is the focus of this study, circulation is dominated by the upper-tropospheric Bolivian High, the lower-level subtropical highs, the Chaco Low over northern Argentina, the \gls{sacz}, and the \gls{sallj} \citep{Grimm2009,Marengo2012}.
Rainfall peaks around \SI{5}{\milli\meter\per\day} during the warm months (October-May) and reaches a minimum near \SI{2}{\milli\meter\per\day} in July and August.
However, the flat topography limits the river's ability to carry the summer runoff, causing seasonal inundation of the Pantanal and distributing the river discharge in time \citep{Bravo2011,Barros2004}.
Thus, the streamflow maxima typically occurs in phase with precipitation upstream of the Pantanal, while downstream of the Pantanal - which we define as the Lower Paraguay River Basin; see \cref{fig:study-area} - the annual peak typically occurs between April and July.

During the warm season (NDJF) a large fraction of rainfall, and nearly all heavy rainfall, in the Paraguay River Basin is associated with mesoscale convection \citep{Velasco1987}.
Previous studies of organized convection and precipitation across subtropical continental South America have found close correspondence to the exit region of the low-level jets \citep{Saulo2007,Salio2007,Marengo2004,Velasco1987}, which is influenced in both summer and winter by midlatitude baroclinic wave trains that interact with the Andes topography to generate orographically bound cyclones and northerly low-level flow \citep{Campetella2002,Seluchi2006,Boers2013,Boers2014}.
The strength and direction of this moisture transport varies substantially between events; \gls{sallj} exit regions range from central Argentina \citep[``Chaco Jet Events'';][]{Salio2002} to Paraguay and southeastern Brazil \citep[``No-Chaco Jet Events'';][]{Vera2006}.

At sub-seasonal timescales, convection in the Paraguay River Basin is modulated by a variety of drivers, notably including the \gls{sacz} and the \gls{mjo}  .
During \gls{sacz} conditions, strong convergence is observed over the Amazon basin with divergence over southwestern Brazil, northern Argentina and Paraguay \citep{Herdies2002,Carvalho2010}; the opposite is true for so-called NoSACZ conditions.
\Gls{sacz} occurrence is related to westerly wind regimes, as well as ``active'' and ``break'' periods of the South American Monsoon System \citep{Marengo2004}.

At seasonal timescales, \gls{enso} is the dominant driver of convection variability in the Paraguay River Basin.
During El Ni\~no years, a low-level anticyclonic anomaly over central Brazil enhances occurrence of the low-level jet, favoring the development of mesoscale convective systems \citep{Velasco1987}.
The intensity and precise extent of this anomaly is relevant for the precise impact of \gls{enso} events.
The region also exhibits substantial variability between seasons of rainfall during El Ni\~no years, including a reversal of rainfall anomalies between November of that year and January of the following one, influenced by land-surface interactions \citep{Grimm2009,Grimm2003}.
Even beyond El Ni\~no years, regional land-surface feedbacks can cause regions that exhibit wet anomalies in the spring to experience more summer precipitation on average \citep{Grimm2007}.
Similarly, mid-latitude dynamics influence low-level wind anomalies on many time scales, though this relationship is complicated due to coupled tropical-extratropical interactions \citep{Carvalho2004,Jones2002}.
To address these potential interactions, a cross-timescale approach based on synoptic circulation types is employed here to diagnose the causes of the rainfall events; this method has been used in previous work for southeastern South America \citep{Munoz2015,Munoz2016} and other regions \citep{Moron2015}.

@@

Rainfall data are taken from the CPC Unified Gauge-based Analysis of Global Daily Precipitation dataset \citep{Xie2010}.
Spatial resolution is \SI{0.5}{\degree}, and temporal resolution is daily.
The retro and real-time datasets are combined and treated as a single dataset.
We define ``heavy'' rainfall events to be exceedances of the \nth{90} percentile; while the value is different for each grid cell, the \nth{90} percentile of area-averaged rainfall over the Lower Paraguay River Basin is approximately \SI{15}{\milli\meter\per\day}.

Atmospheric circulations are diagnosed using daily data from the NCAR-NCEP Reanalysis II dataset \citep{Kanamitsu2002}.
Spatial resolution is \SI{2.5}{\degree}.
Because the end-of-day time for the rainfall data is 12:00 GMT over most of South America \citep{Xie2010}, we use six-hour reanalysis data, and shift by twelve hours before re-sampling to the daily time step.
This ensures that the time steps in the reanalysis and rainfall data sets are the same, but means that a day is defined as beginning at 12:00 GMT.
Since most summer rainfall in this region occurs overnight, this end-of-day time (which translates to approximately 8:00 AM locally depending on the exact time zone) tends to separate distinct events.
Oceanic \gls{sst} patterns are explored at the monthly time step using the \SI{1}{\degree} NOAA OI.v2 dataset \citep{Reynolds2002}.

Streamflow data was collected by the Paraguayan Navy and National Administration of Navigation and Ports of Paraguay, and was processed and distributed by the Paraguayan Directorate of Meteorology and Hydrology.
Locations of streamflow gauges are shown in \cref{fig:study-area}.
Because no stage-discharge curves are available, we present only the river stage values; while this is relevant from the perspective of flood damage, flow rates cannot be estimated without these curves (which are difficult to reconstruct as river geometry changes over time).

This study also makes use of some climate indices.
Data on \gls{enso}, specifically the NI\~NO3.4 index, came from a statistical-dynamical interpolation \citep{Kaplan1998}, which is constrained by relatively high-quality observations during the study period.
Data on the \gls{mjo}   came from the Australian Bureau of Meteorology \citep{Wheeler2004}.

@@

The seasonal predictions used were ``flexible format'' forecasts provided by the \gls{iri}.
This product provides the probability of exceedance (or non-exceedance) of a user-specified percentile of the climatological distribution.
The DJF 2015-2016 forecasts analyzed were produced in November 2015.
Due to the short record of flexible format forecasts (2012-2016), no verification was performed for seasonal predictions.
These forecasts are provided at a horizontal resolution of \SI{2.5}{\degree}.

The sub-seasonal forecasts used were issued by the \gls{ecmwf} using a coupled model, available via the \gls{s2s} Prediction Project Database \citep{Vitart2016} at \SI{1.5}{\degree} resolution.
Forecasts consider the period starting in Dec 2015 until Mar 2016 and hindcasts to assess the real-time predictive skill consider the period Dec 1-7, 1995-2014.
There is a total of 51 ensemble members for each forecast, and 11 ensemble members for each of the 20 hindcasts (Dec 1-7, 1995-2014).

@@

Anomalies were calculated relative to the seasonal mean from November 1979 to February 2016, and the anomalies thus contain information on intra-seasonal variability.

@@

Several types of analyses are used to diagnose the causes of the heavy rainfall events, and to bias-correct and verify the forecasts.
Computation was performed in the python environment using stable open source packages \citep{Hoyer2017,vanderWalt2011,McKinney2010,Hunter2007}.
All codes to reproduce or modify this analysis are available at \url{https://github.com/jdossgollin/PYFloods}.

Given the behavior of the Paraguay River discussed above, we define the Lower Paraguay River Basin as the region bounded by \SIrange{-59.75}{-55.75}{\degree W} and \SIrange{26.75}{22.75}{\degree S}, as shown in \cref{fig:study-area}.
In this region, given topography and previous studies \citep{Barros2004,Bravo2011}, one might hypothesize rainfall inputs to most closely correspond to river levels at the stream gauges in \cref{fig:study-area}.
Results are not sensitive to small variations in this domain.

@@

The clustering was performed on the daily NDJF \SI{850}{\hecto\pascal} streamfunction field ($\psi$), calculated by integrating the meridional and zonal wind fields using spherical harmonics, as implemented in the \texttt{windspharm} package \citep{Dawson2016}, over the domain spanning \SIrange{15}{30}{\degree S} and \SIrange{65}{45}{\degree W} (\cref{fig:study-area}).

To facilitate clustering (which tends to perform poorly in high-dimensional spaces), the NDJF anomaly field of $\psi_{850}$ was projected  onto its four leading \glspl{eof}, accounting for $>95\%$ of the total observed variance.
No meridional weighting was applied as the selected domain is relatively small and does not extend into high latitudes.
Once the \glspl{eof} were calculated, the principal component time series were computed for each day and scaled to unit variance.
This rescaling is not a necessary step; its effect is to treat all retained principal components as equally important, which provides relatively greater weight to \glspl{eof} 2, 3, and 4 than carrying out the clustering without re-scaling.
Though our approach of first selecting the number of \glspl{eof} to use and then choosing to scale them equally involves more subjective decisions than an approach without rescaling, in this case the resulting physical patterns represented by the \glspl{eof} more closely represent patterns identified in the literature; this is further discussed in \cref{sec:diagnostics}.

Next, the $K$-means algorithm was used to assign a single cluster value to each day on record using the 4-dimension principal component time series.
The $K$-means technique is a partitioning method that classifies all days in the study into a predefined number of clusters.
The algorithm proceeds as follows:
\begin{enumerate}
	\item Randomly choose $K$ cluster centers $\mu_1^{(0)}, \ldots, \mu_K^{(0)}$ (where $0$ refers to the $0$th iteration)
	\item Iterate until convergence, indexing each iteration with $j$:
	\begin{enumerate}
		\item Assign each observation (day) $x_i$ to the nearest cluster center; we define this using the Euclidean distance but other measures, such as the Mahalanobis distance, could also be used:
		\begin{equation}
			m_i^{(j+1)} := \arg\min_{k\in 1,\ldots,K} \abs{\abs{x_i - \mu_k^{(j)}}}
		\end{equation}
		\item Recompute the cluster centers as the mean of all points assigned to that cluster
		\begin{equation}
			\mu_k^{(j+1)} := \frac{1}{\abs{\qty{i \big| m_i^{(j+1)}=k}}} \sum_{i | m_i^{(j+1)}=k} x_i
		\end{equation}
		where $\abs{\cdot}$ denotes vector length.
		\item Stop iteration if the change in centroids $\mu^{(j+1)} - \mu^{(j)}$ is less than a small but non-zero tolerance parameter $\tau$.
	\end{enumerate}
\end{enumerate}
The cluster centroids $\mu_k$ produced by the $K$-means algorithm can then be interpreted as a Voronoi decomposition of the phase space into $K$ regions, and specifically as the Voronoi diagram which minimizes within-cluster variance.

The $K$-means algorithm is guaranteed to converge to a local minimum of intercluster variance; to select the best partition, 500 simulations were created using the implementation in Python's \texttt{scikit-learn} package \citep{Pedregosa2012} and then the classifiability index of \citet{Michelangeli1995} was computed between each partition and the 499 others.
The partition whose classifiability index, averaged for all 499 pairwise comparisons, was the highest was selected.
Calculation of the classifiability index for several values of $K$ (supplemental figure S3)\footnote{CHECK} suggests that states with $K=5,6,\ldots,9$ are all reasonable; we chose the solution $K=6$ because the clusters identified are qualitatively similar to those determined over southeastern South America \citep{Munoz2015,Munoz2016} and have an intuitive physical meaning, discussed further in the following sections.
We refer to the resulting clusters as \glspl{wt}.

\subsection{Forecasts and Model Output Statistics}

A wide variety of methods, generically known as \gls{mos} \citep{Glahn1972}, have been proposed to correct for different types of bias in model outputs.
In this work, we analyze how well the rainfall events could have been predicted, both using the raw sub-seasonal forecasts and  MOS-adjusted sub-seasonal forecasts.
We use  four types of \gls{mos} techniques: the homoscedastic extended logistic regression \gls{xlr}; the \gls{hxlr}; \gls{pcr}; and \gls{cca}.

Logistic regression models the conditional probability of binary events, and it has been widely used in \gls{mos} approaches for binary predictands \citep{Hamill2004}.
Nonetheless, when using the logistic regression to address multiple thresholds via independent fits, the predicted probabilities are, in general, not mutually consistent \citep{Messner2014}.
The \gls{xlr} was designed to address this shortcoming \citep{Wilks2009}, via the consideration of a transformation of the thresholds of interest as an additional predictor variable.
The \gls{hxlr}, a heteroscedastic generalization of the \gls{xlr}, was proposed to appropriately use the ensemble spread as predictor for the dispersion of the predictive distribution \citep{Messner2014}.

\Gls{cca} is a common statistical method frequently used to forecast rainfall using a purely empirical approach \citep{Mason2008,Barnston2012,Jolliffe2012,Barnston1992,Wilks2006}.
\Gls{cca} identifies modes of co-variability, called canonical variates or canonical modes, by maximizing the correlation between linear combinations of the predictor and predictand's \glspl{eof}.
The method forecasts spatial patterns of variability spanning across the region of interest rather than making forecasts for individual locations.
In \gls{pcr}, a special case of \gls{cca}, each grid cell in the predictand field is estimated by regression using a linear combination of the predictor's \glspl{eof} \citep{Mason2008,Wilks2006} rather than by identifying canonical modes.
Unlike the \gls{xlr} and \gls{hxlr} models, which perform bias correction independently for each grid cell, the \gls{cca} and \gls{pcr} models can address biases in both the magnitude and the spatial distribution of the modeled precipitation patterns.

For the purposes of \gls{mos} corrections, the predictand (variable to forecast) is the observed rainfall for the target period of interest, and the predictor (variable to be corrected) is the raw model rainfall for the same period.
Exceedance of the \nth{90} percentile during the 1995-2014 period is used to define the heavy event cases.
We use the same spatial domain [\SIrange{39}{17}{\degree S}; \SIrange{66}{49}{\degree W}] for both the predictor and the predictand, except for the \gls{pcr} and \gls{cca} cases, in which a larger domain [\SIrange{0}{60}{\degree S}; \SIrange{80}{30}{\degree W}] was used to better capture the spatial patterns in the raw model field.
A variety of domains and ways to combine initialization times were explored; the best results were selected in terms of the corresponding Kendall's $\tau$ rank correlation coefficient.
A summary of the final candidate predictors found to be most skillful for each \gls{mos} model is presented in \cref{tab:mos-methods}.

@@

The \gls{iri}'s seasonal predictions are already provided with bias correction \citep{Barnston2010}, and thus we did not perform any further \gls{mos} on seasonal rainfall fields.

@@

In addition to visually comparing predictions and observations to verify how well the heavy rainfall events could have been predicted, we use the Ignorance Score,
\begin{equation}\label{eq:ignorance}
	\text{IGN} \equiv - \log_2 p(Y),
\end{equation}
where $Y$ is the observed outcome and $p(Y)$ is the density function of the forecast distribution \citep{Good1952,Roulston2002,Brocker2007}.
The Ignorance Score was introduced as an information theory-based verification measure,  decomposable into easily interpretable components: reliability, resolution and uncertainty \citep{Weijs2010}.
Due to its close relationship to Shannon's information entropy, it is used to measure forecast utility, or the amount of information gain expected from a forecast \citep{Roulston2002}.

We also compute the Generalized Relative Operating Characteristics score, commonly referred to as the 2AFC score \citep{Mason2009}, to evaluate skill of probabilistic rainfall forecasts.
This score measures the ``proportion of all available pairs of observations of differing category whose probability forecasts are discriminated in the correct direction'' \citep{Mason2009}.
It has an intuitive interpretation as an indication of how often the forecasts are correct.
